Supplementary Document

### 2019-03-19 paper/main.tex
adding supplements template files

\newif\ifpreprint
%\preprinttrue

@@

\begin{abstract}
\noindent\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. We showed that these classifiers outperformed the other methods with the best classifiers ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
\\
\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\end{abstract}

@@

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. % to show how the classifiers tend to outperform the other methods.  
We showed that these classifiers outperformed the other methods with the best classifier ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
%this was probably too detailed, need a different sentence
%The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  
\\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Given a prioritized list of variants, analysts manually inspect those variants and curate a list of variants to ultimately report to the ordering physician.  Unfortunately, manual curation is a time consuming process where analysts must inspect each variant while maintaining a mental picture of the patient's phenotype.  \citealp{bick2017} report an average of 600 variants per case analyzed by two people (one analyst and one director) over three hours ($\approx$100 variants per man-hour).  If causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.  Finally, accurate prioritization is a step towards the ultimate goal of automatic identification of rare variants that may cause a patient's primary phenotypes.

One of the issues with previously published ranking methods is that they were primarily tested on simulated datasets with known, single-gene, pathogenic variants injected into real or simulated background genomic datasets.  Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few (typically 2-3) noisy terms added or removed.  Unfortunately, rare disease patient data is often far noisier due to a wide variety of reasons.

In this paper, we focus on real patient data from the multi-site collaboration of the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.  Patients accepted into the UDN are believed to have rare, undiagnosed diseases of genetic origin.  Because the UDN is not focused on a single particular disease, the patient population has a diverse range of phenotypes represented.  Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as phenotypes caused by old age).  Because the UDN is a research collaboration, there is also variability in reported variants that range in pathogenicity from ``variant of uncertain significance" (VUS) through ``pathogenic" as defined by the ACMG guidelines \citep{richards2015}. The summation of this real-world variation means that accurately prioritizing variants is challenging due to noise and variation in phenotype inputs and variation in pathogenicity of reported variant outputs.

@@

Given the above variant information, we split the data into training and testing sets for measuring the performance of classifiers to predict whether a variant would be clinically reported or not.  We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.  Of note, our focus was not on picking the ``best'' classifier, but rather on analyzing their overall ability of classification algorithms to handle the noise of real-world patient cases from the UDN.

Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing us to measure their performance as a classifier and as a ranking system.  After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant ranking perspective.  Additionally, we tested four publicly available variant ranking algorithms and two single-value ranking methods for comparison.  All of the scripts to train classifiers, test classifiers, and format results are contained in the VarSight repository.

@@

\else
\begin{methods}
\fi
\section{Methods}
\subsection{Data sources}
All samples were selected from the cohort of Undiagnosed Diseases Network (UDN) \citep{ramoni2017} genome sequencing samples that were originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB).  In short, the UDN accepts patients with rare, undiagnosed diseases that are believed to have a genetic origin.  The UDN is not restricted to a particular disease, so there are a diverse set of diseases and phenotypes represented across the whole population.  The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) some patients have multiple diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, and 3) some patients exhibit more or fewer phenotypes than are associated with the classic disease presentation.  For more details on the UDN, refer to \citealp{ramoni2017}.

@@

The first method uses phenotype-to-gene annotations provided by the HPO to calculate a cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  In short, this method builds a vector from the phenotype terms (and ancestor terms in the ontology) for both the patient and each gene.  Then, the cosine of the angle between the patient vector and each gene vector is calculated as a representation of similarity.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.

@@

We required the following for each proband sample included in our analyses: 1) at least one clinically reported primary variant that came through the primary clinical filter (i.e. it was not found through some other targeted search) and 2) a set of phenotypes annotated with Human Phenotype Ontology \citep{kohler2018} terms using the Phenotips software \citep{girdea2013}.  At the time of writing, this amounted to 378 primary, reported variants spanning a total of 237 proband samples.

\subsection{Feature Selection}
\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}
\caption{Feature selection.  This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.  After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).  We average the two importance values, and order them from most to least important.  Feature labels with an `\_' represent a single category of a multi-category feature (i.e. ``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem).}
\label{tab:features}
\end{table}

@@

\begin{table}[!t]
\processtable{Feature selection.  This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.  After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).  We average the two importance values, and order them from most to least important.  Feature labels with an `\_' represent a single category of a multi-category feature (i.e. ``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem). \label{tab:features}}{
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

For the purposes of classification, all annotations needed to be cleaned and stored as numerical features.  For single-value numerical annotations (e.g. float values like CADD or GERP), we simply copied the annotation over as a single value feature.  Missing annotations were assigned a default value that was outside the expected value range for that feature.  Additionally, these default values were always on the less impactful side of the spectrum (e.g. a default conservation score would err on the side of not being conserved).  The one exception to this rule was for variant allele frequencies where a variant absent from the database was considered to have an allele frequency of 0.0.  For multi-value numerical annotations, we reduced the values (using minimum or maximum) to a single value that was used as the feature.

@@

After converting all annotations to numerical features or multiple bin-count encoded features, we had a total of 95 features per variant.  We then pruned down to only the top 20 features using univariate feature selection (specifically the \texttt{SelectKBest} method of \texttt{sklearn}).  This method evaluates how well an individual feature performs as a classifier and keeps only the top 20 features for the full classifiers.  Table \ref{tab:features} shows the list of retained features ordered by feature importance after training.  Feature importance was derived from the random forest models which automatically report how important each feature was to classification.  The entire set of annotations along with descriptions of how each was processed prior to feature selection are detailed in the Supplementary Documents.

\subsection{Model training and tuning}
As noted earlier, there are generally hundreds of variants per proband that pass the filter, but only a few are ever clinically reported.  Across all 237 proband samples, there were a total of 378 clinically reported variants and another 87819 variants that were seen but not reported.  As a result, there is a major imbalance in the number of true positives (variants clinically reported) and true negatives (variants seen, but not clinically reported).

We split the data into training and test sets on a per-proband basis with the primary goal of roughly balancing the total number of true positives in each set.  Additionally, the cases were assigned to a particular set by chronological order of analysis in order to reduce any chronological biases that may be introduced by expanding scientific knowledge (i.e. there are roughly equal proportions of ``early" or ``late" proband samples from the UDN in each set).  In the training set, there were a total of 189 returned variants and 44593 not returned variants spanning 120 different probands.  In the test set, there were a total of 189 returned variants and 43226 not returned variants spanning 117 different probands.  In our results, the returned test variants are further stratified in their reported levels of pathogenicity.

We then selected four publicly available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.  These classifiers were chosen for three main reasons: 1) their ability to handle imbalanced data (i.e. far more unreported variants than reported variants), 2) their ability to scale to the size of the training and testing datasets, and 3) they are freely available implementations that can be tuned and tested with relative ease.  The two random forest models build collections of decision trees that weight each training input by its class frequency.  Logistic regression calculates the probability of a value belonging to a particular class, again weighting by the class frequency.  In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.  For more details on each classifier, please refer to the {\it sklearn} and {\it imblearn} documentations.

Initially, we also tested the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.  Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.  While we did not achieve positive results using these three implementations, it may be possible to use the methods through another implementation.

For each of our tested models, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each model and set of hyperparameters, we performed stratified 10-fold cross validation on the training variants and recorded the balanced accuracy (i.e. weighted accuracy based on inverse class frequency) and the F1 scores (i.e. harmonic mean between precision and recall).  For each model type, we saved the hyperparameters and model with the best average F1 score.  These four tuned models were then trained on the full training set and tested against the unseen set of test proband cases.  The set of hyperparameters tested along with the highest performance setting for each hyperparameter can be found in the Supplementary Documents.

@@

\section{Results}%mana said results!
\subsection{Classifier Statistics}
\begin{figure*}
\centerline{
\includegraphics[width=0.45\textwidth]{codi_rf_roc.png}
\includegraphics[width=0.45\textwidth]{codi_rf_pr.png}
}
\caption{Receiver operator and precision-recall curves.  
These figures show the performance of the four classifiers on the testing set after hyperparameter tuning and fitting to the training set.  On the left, we show the receiver operator curve (false positive rate against the true positive rate).  On the right, we show the precision recall curve.  Area under the curve (AUROC and AUPRC) is reported beside each method in the legend.}
\label{fig:roc}
\end{figure*}

Each model was tuned using 10-fold cross validation and the resulting average and standard deviation of balanced accuracy is reported in Table \ref{tab:classifier_results}.  After fitting the tuned model to the full training set, we evaluated the models on the testing set by calculating the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC).  Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the results from the testing set on all four models.

From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  However, these classifier all have relatively poor performance from a precision-recall perspective (best AUPRC was 0.2109).  This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

@@

In addition to the model performance statistics, we also quantified the performance of each classifier as a ranking system.  For each proband, we calculated the probability of each class (reported or not reported) for each variant and ordered them from highest to lowest probability of being reported.  We then calculated median and mean rank statistics for the reported variants.  Additionally, we quantified the percentage of reported variants that were ranked in the top 1, 10, and 20 variants in each case.  While the models were trained as a binary classification system, we broke down the results further to demonstrate differences between variants that were clinically reported as a variant of uncertain significance (VUS), likely pathogenic, and pathogenic.

@@

%\subsection{Random Forest Feature Importances}
%After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.  Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).  The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.  By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.  Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.

@@

\section{Conclusion}
We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report for rare disease patients.  We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process obtained from the Undiagnosed Diseases Network (UDN).  From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.  However, when evaluated as a ranking system, all four methods out-performed the single-measure ranking systems and external tools that were tested.  The classifiers had median ranks of 6.0-10.0 for all reported variants while ranking 65-72\% of those variants in the top 20 for the case.  For ``Pathogenic" variants, the median ranks were 1.0-4.0 and 80-94\% of those variants were ranked in the top 20 for the case.  While these algorithms are not perfect classifiers, their use as a prioritization system is quite promising.

There are two major factors that we believe is influencing the classifiers' performance relative to the externally tested tools.  First, all results were generated using real-world patients from the UDN, but only our four classifiers were trained on real-world patients from the UDN.  In contrast, the four external tools were primarily evaluated and/or trained using simulations that do not capture the variation and/or noise that is apparent in UDN patients.  Second, the four classifiers we tested have far more information (i.e. features) available to them than the external tools.  As noted in our methods, we tried to reflect an analyst's view of each variant as much as possible, ultimately resulting in 50 features incorporated into the model.  Incorporating the same set of features and/or training on real-world patients may improve the externally tested tools with respect to these classifiers.

@@

Overall, we believe the classifiers trained in VarSight represent a significant step forward in tackling real clinical data.  The tested models improved our ability to prioritize variants despite the variability and uncertainty injected by real-world patients.  Ultimately, we believe implementing these classifiers will enable analysts to assess the best candidate variants first, reducing the time to analyze a case and return molecular diagnoses to patients.

>>>>

\newif\ifpreprint
\preprinttrue

@@

\begin{abstract}
\noindent\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant annotations along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to four variant prioritization algorithms and two single-measure controls. We showed that these classifiers outperformed the other methods with the best classifiers ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
\\
\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight} release v1.1. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\end{abstract}

@@

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant annotations along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to four variant prioritization algorithms and two single-measure controls. % to show how the classifiers tend to outperform the other methods.  
We showed that these classifiers outperformed the other methods with the best classifier ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
%this was probably too detailed, need a different sentence
%The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  
\\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight} release v1.1. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Given a prioritized list of variants, analysts manually inspect those variants and curate a list of variants to ultimately report to the ordering physician.  Unfortunately, manual curation is a time consuming process where analysts must inspect each variant while maintaining a mental picture of the patient's phenotype.  \citealp{bick2017} report an average of 600 variants per case analyzed by two people (one analyst and one director) over three hours, meaning a thoughput of $\approx$100 variants per man-hour.  If causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited, reducing the total number of variants reviewed and therefor the time to analyze a case.  Additionally, accurate prioritization is a step towards the ultimate goal of automatic identification of rare variants that may cause a patient's primary phenotypes.

One of the issues with previously published ranking methods is that they were primarily tested on simulated datasets with known, single-gene, pathogenic variants injected into real or simulated background genomic datasets.  Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few (typically 2-3) noisy terms added or removed.  In practice, rare disease patient data is often far noisier due to a wide variety of reasons.% such as multiple diseases, incomplete penetrant phenotypes, or age-related phenotypes not caused by the primary disease.

In this paper, we focus on real patient data from the multi-site collaboration of the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.  Patients accepted into the UDN are believed to have rare, undiagnosed diseases of genetic origin.  Because the UDN is not focused on a single particular disease, the patient population has a diverse range of phenotypes represented.  Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as age- or pathogen-related disease).  Because the UDN is a research collaboration, there is also variability in reported variants that range in pathogenicity from ``variant of uncertain significance" (VUS) through ``pathogenic" as defined by the ACMG guidelines \citep{richards2015}. The summation of this real-world variation means that accurately prioritizing variants is challenging due to noise and variation in phenotype inputs and variation in pathogenicity of reported variant outputs.

@@

Given the above variant information, we split the data into training and testing sets for measuring the performance of classifiers to predict whether a variant would be clinically reported or not.  We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.  Of note, our focus was not on picking the ``best'' classifier, but rather on analyzing their overall ability to handle the noise of real-world patient cases from the UDN.

Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing us to measure their performance as both a classifier and a prioritization/ranking system.  After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant prioritization perspective.  Additionally, we tested four publicly available variant prioritization algorithms and two single-value ranking methods for comparison.  All of the scripts to train classifiers, test classifiers, and format results are contained in the VarSight repository.

@@

\else
\begin{methods}
\fi
\section{Methods}
\subsection{Data sources}
All samples were selected from the cohort of Undiagnosed Diseases Network (UDN) \citep{ramoni2017} genome sequencing samples that were sequenced at HudsonAlpha Institute for Biotechnology.  In short, the UDN accepts patients with rare, undiagnosed diseases that are believed to have a genetic origin.  The UDN is not restricted to a particular disease, so there are a diverse set of diseases and phenotypes represented across the whole population.  The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) patients may have multiple genetic diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, 3) patients may exhibit more or fewer phenotypes than are associated with the classic disease presentation, and 4) patients may have phenotypes of non-genetic origin such as age- or pathogen-related phenotypes.  For more details on the UDN, refer to \citealp{ramoni2017}.

@@

The first method uses phenotype-to-gene annotations provided by the HPO to calculate a cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  Given $P$ terms in the HPO, this method builds a binary, $P$-dimensional vector for each patient such that only the phenotype terms and all ancestral phenotype terms tied to the patient are set to 1.  Similarly, a $P$-dimensional vector for each gene is built using the phenotype-to-gene annotations.  Then, the cosine of the angle between the patient vector and each gene vector is calculated as a representation of similarity.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.

@@

We required the following for each proband sample included in our analyses: 1) at least one clinically reported primary variant that came through the primary clinical filter (i.e. it was not found through some other targeted search) and 2) a set of phenotypes annotated with Human Phenotype Ontology \citep{kohler2018} terms using the Phenotips software \citep{girdea2013}.  At the time of writing, this amounted to 378 primary-reported variants and 87819 unreported variants spanning a total of 237 proband samples.

\subsection{Feature Selection}
\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}
\caption{Feature selection.  This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.  After training, the two random forest classifiers report the importance of each feature in the classifier (total is 1.00 per classifier).  We average the two importance values, and order them from most to least important.  Feature labels with an `\_' represent a single category of a multi-category feature (i.e. ``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem).}
\label{tab:features}
\end{table}

@@

\begin{table}[!t]
\processtable{Feature selection.  This table shows the 20 features gathered using univariate feature selection that were used to train the classifiers.  After training, the two random forest classifiers report the importance of each feature in the classifier (total is 1.00 per classifier).  We average the two importance values, and order them from most to least important according to this average.  Feature labels with an `\_' represent a single category of a multi-category feature (i.e. ``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem). \label{tab:features}}{
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

For the purposes of classification, all annotations needed to be cleaned and stored as numerical features.  For single-value numerical annotations (e.g. float values like CADD or GERP), we simply copied the annotation over as a single value feature.  Missing annotations were assigned a default value that was outside the expected value range for that feature.  Additionally, these default values were always on the less impactful side of the spectrum (e.g. a default conservation score would err on the side of not being conserved).  The one exception to this rule was for variant allele frequencies where a variant absent from the database was considered to have an allele frequency of 0.0.  For multi-value numerical annotations, we reduced the values (using minimum or maximum) to a single value corresponding to the ``worst'' value (i.e. most deleterious value, most conserved value, etc.) that was used as the feature.

@@

After converting all annotations to numerical features or multiple bin-count encoded features, we had a total of 95 features per variant.  We then pruned down to only the top 20 features using univariate feature selection (specifically the \texttt{SelectKBest} method of \textit{sklearn} \citep{pedregosa2011}).  This method evaluates how well an individual feature performs as a classifier and keeps only the top 20 features for the full classifiers.  Table \ref{tab:features} shows the list of retained features ordered by feature importance after training.  Feature importance was derived from the random forest classifiers which automatically report how important each feature was for classification.  The entire set of annotations along with descriptions of how each was processed prior to feature selection are detailed in the Supplementary Documents.

\subsection{Classifier training and tuning}
As noted earlier, there are generally hundreds of variants per proband that pass the filter, but only a few are ever clinically reported.  Across all 237 proband samples, there were a total of 378 clinically reported variants and another 87819 variants that were seen but not reported.  As a result, there is a major imbalance in the number of true positives (variants clinically reported) and true negatives (variants seen, but not clinically reported).

We split the data into training and test sets on a per-proband basis with the primary goal of roughly balancing the total number of true positives in each set.  Additionally, the cases were assigned to a particular set by chronological order of analysis in order to reduce any chronological biases that may be introduced by expanding scientific knowledge (i.e. there are roughly equal proportions of ``early" or ``late" proband samples from the UDN in each set).  In the training set, there were a total of 189 returned variants and 44593 not returned variants spanning 120 different probands.  In the test set, there were a total of 189 returned variants and 43226 not returned variants spanning 117 different probands.  In our results, the returned test variants are further stratified by their reported levels of pathogenicity.

We then selected four publicly available binary-classification models that are capable of training on imbalanced datasets: the \texttt{RandomForest} model by {\it sklearn} \citep{pedregosa2011}, the \texttt{LogisticRegression} model by {\it sklearn}, the \texttt{BalancedRandomForest} model by {\it imblearn} \citep{lemaitre2017}, and the \texttt{EasyEnsembleClassifier} model by {\it imblearn}.  These classifiers were chosen for three main reasons: 1) their ability to handle imbalanced data (i.e. far more unreported variants than reported variants), 2) their ability to scale to the size of the training and testing datasets, and 3) they are freely available implementations that can be tuned, trained, and tested with relative ease in the same Python framework.  The two random forest classifiers build collections of decision trees that weight each training input by its class frequency.  Logistic regression calculates the probability of a value belonging to a particular class, again weighting by the class frequency.  In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.  For more details on each classifier, please refer to the {\it sklearn} and {\it imblearn} documentations \citep{pedregosa2011, lemaitre2017}.

Initially, we also tested the support vector classifier by {\it sklearn} (\texttt{SVC}), the multi-layer perceptron by {\it sklearn} (\texttt{MLPClassifier}), and the random under-sampling AdaBoost classifier by {\it imblearn} (\texttt{RUSBoostClassifier}).  Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.  While we did not achieve positive results using these three implementations, it may be possible to use the methods through another implementation.

For each of our tested classifiers, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each classifier and set of hyperparameters, we performed stratified 10-fold cross validation on the training variants and recorded the balanced accuracy (i.e. weighted accuracy based on inverse class frequency) and the F1 scores (i.e. harmonic mean between precision and recall).  For each classifier type, we saved the hyperparameters and classifier with the best average F1 score (this is recommended for imbalanced datasets).  These four tuned classifiers were then trained on the full training set and tested against the unseen set of test proband cases.  The set of hyperparameters tested along with the highest performance setting for each hyperparameter can be found in the Supplementary Documents.

@@

\section{Results}%mana said results!
\subsection{Classifier Statistics}
\begin{figure*}
\centerline{
\includegraphics[width=0.45\textwidth]{codi_rf_roc.png}
\includegraphics[width=0.45\textwidth]{codi_rf_pr.png}
}
\caption{Receiver operator and precision-recall curves.  
These figures show the performance of the four classifiers on the testing set after hyperparameter tuning and fitting to the training set.  On the left, we show the receiver operator curve (false positive rate against the true positive rate).  On the right, we show the precision recall curve.  Area under the curve (AUROC or AUPRC) is reported beside each method in the legend.}
\label{fig:roc}
\end{figure*}

The hyperparameters for each classifier were tuned using 10-fold cross validation and the resulting average and standard deviation of balanced accuracy is reported in Table \ref{tab:classifier_results}.  After fitting the tuned classifiers to the full training set, we evaluated the classifiers on the testing set by calculating the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC) (also shown in Table \ref{tab:classifier_results}).  Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the results from the testing set on all four classifiers.

From these metrics, we can see that all four classifiers have a similar performance with regards to AUROC.  However, all classifiers have a relatively poor performance from a precision-recall perspective (best AUPRC was only 0.2458).  This indicates that from a classification perspective, these classifiers would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

@@

In addition to the classifier performance statistics, we also quantified the performance of each classifier as a ranking system.  For each proband, we calculated the probability of each class (reported or not reported) for each variant and ordered them from highest to lowest probability of being reported.  We then calculated median and mean rank statistics for the reported variants.  Additionally, we quantified the percentage of reported variants that were ranked in the top 1, 10, and 20 variants in each case.  While the classifiers were trained as a binary classification system, we broke down the results further to demonstrate differences between variants that were clinically reported as a variant of uncertain significance (VUS), likely pathogenic, and pathogenic.

@@

%\subsection{Random Forest Feature Importances}
%After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.  Both random forest classifiers calculate a feature importance array that stores how important each feature is in the trained classifier (the other two classifiers do not have this information readily available).  The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained classifier.  By chance, we had exactly 50 parameters as input to our classifier, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.  Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.

@@

\section{Conclusion}
We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report for rare disease patients.  We trained and tested these algorithms using real patient variants and phenotype terms obtained from the Undiagnosed Diseases Network (UDN).  From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.  However, when evaluated as a ranking system, all four methods out-performed the single-measure ranking systems and external tools that were tested.  The classifiers had median ranks of 6.0-10.0 for all reported variants and ranked 65-72\% of those variants in the top 20 for the case.  For ``Pathogenic" variants, the median ranks were 1.0-4.0 and 80-94\% of those variants were ranked in the top 20 for the case.  While these algorithms are not perfect classifiers, their use as a prioritization system is quite promising.

There are two major factors that we believe is influencing the classifiers' performance relative to the externally tested tools.  First, all results were generated using real-world patients from the UDN, but only our four classifiers were trained on real-world patients from the UDN.  In contrast, the four external tools were primarily evaluated and/or trained using simulations that do not capture the variation and/or noise that is apparent in UDN patients.  Second, the four classifiers we tested have far more information (i.e. features) available to them than the external tools.  As noted in our methods, we tried to reflect an analyst's view of each variant as much as possible, starting with 95 features that were pruned down to 20 features used by each classifier.  Incorporating the same set of features and/or training on real-world patients may improve the externally tested tools with respect to these classifiers.

@@

Overall, we believe the classifiers trained in VarSight represent a significant step forward in tackling real clinical data.  The tested classifiers improved our ability to prioritize variants despite the variability and uncertainty injected by real-world patients.  Ultimately, we believe implementing these classifiers will enable analysts to assess the best candidate variants first, allowing for faster clinical throughput and increased automation in the future.

### 2019-03-19 paper/supplements/supplement.tex
adding supplements template files

\subsubsection{HPO-cosine}
The first phenotype feature is called the ``HPO-cosine'' score.  Given $N$ terms in the HPO, a phenotype profile is represented by an $N$-dimensional vector where dimension $p$ is set to 1.0 if the corresponding phenotype term is present in the profile.  Note that if a phenotype is present in the profile, all ancestor terms as defined by the ontology are automatically set as present as well.  Given two profiles, we calculate the ``HPO-cosine'' score by performing the dot product calculation on the vectors and normalizing by their lengths.  Using this measure, values close to 1.0 are highly similar (the vectors are ``pointing'' in similar directions) whereas those close to 0.0 are not (the vectors are mostly perpendicular).

>>>>

\subsubsection{HPO-cosine}
The first phenotype feature is called the ``HPO-cosine'' score.  Given $P$ terms in the HPO, a phenotype profile is represented by an $P$-dimensional vector where dimension $p$ is set to 1.0 if the corresponding phenotype term is present in the profile.  Note that if a phenotype is present in the profile, all ancestor terms as defined by the ontology are automatically set as present as well.  Given two profiles, we calculate the ``HPO-cosine'' score by performing the dot product calculation on the vectors and normalizing by their lengths.  Using this measure, values close to 1.0 are highly similar (the vectors are ``pointing'' in similar directions) whereas those close to 0.0 are not (the vectors are mostly perpendicular).
